A comps-based approach for interpreting tree-based predictions with an application to the NFL draft

### 2025-07-26 articles/jsa/paper.tex
upload JSA file

\documentclass{article}

@@

All data used for this study are publicly available from Sports Reference (\cite{sports_reference_sports_2025}). Our dataset includes 2,099 quarterbacks who played their final season at the NCAA Division I Football Bowl Subdivision (FBS) between 2001 and 2023 (inclusive) who played more than six total games in their careers and attempted more than five passes per season. The data for each player-season include team, conference, strength of schedule, games played, passing statistics (attempts, completions, touchdowns, interceptions), rushing statistics (attempts, yards, touchdowns), and awards (All-America designation, Heisman voting).

@@

Another limitation of average QBR per season is that some players put up very high QBR numbers in very small samples (e.g. a QBR of 64.7 on only 26 passing attempts over a full season) and are rated as having very positive outcomes although a more subjective evaluation of their careers would be less positive. To mitigate the effect of this phenomenon, we regress each player's career QBR per season toward zero:
\begin{equation*}
  \mbox{QBR}_{\mbox{reg}} = \frac{(\mbox{Attempts / Season}) \cdot (\mbox{Average QBR}) + c \cdot 0}{(\mbox{Attempts / Season}) + c}.
\end{equation*}
We chose $c = 20.8$ as this value minimizes the error in predicting a player's next-season QBR when used to regrees current-season QBR.

A response variable such as nflWAR (\cite{yurko_nflwar_2019}) or Sports Reference's Approximate Value could have avoided some of the limitations of QBR, but we were unable to obtain these data for a sufficient sample of players' careers. The primary purpose of the present work (extracting similarity scores to interpret random forest predictions through the lens of player comps) holds regardless of the response variable used.

@@

To train the random forest model, we first implemented a random 70\%/30\% train/test split. We chose $B$ = 1,000 decision trees, each trained on a bootstramp sample of the data, with replacement. To determine the number of variables to consider at each node and the maximum tree size, we used the R package tuneRanger (\cite{probst_tuneranger_2018}), which uses model-based optimization of out-of-bag prediction error to tune the hyperparameters. After 100 iterations of the tuning algorithm, we estimated the optimal number of variables to consider at each node to be 5 and the optimal minimum node size to be 73.

@@

Figure \ref{fig:importance} shows the importance of each feature learned by the random forest model. The model's test RMSE was 8.61 units of QBR, corresponding to explaining 43.7\% of the variance of QBR in the test set. Final-season Heisman voting, average passing performance (yards, touchdowns, completions) per season, and final-season strength of schedule stand out as the most important predictors of NFL success. Note that yards per season, touchdowns per season, and completions per season are highly correlated with each other---each pairwise correlation is at least 0.94.

@@

\begin{figure}
    \centering
    \includegraphics[width=0.75\linewidth]{../../figures/3d_plot.pdf}
    \caption{\textit{Predicted adjusted QBR plotted against average passing yards per season and final-season passer rating.}}
    \label{fig:3d-plot}
\end{figure}

@@

\begin{table}[H]
  \centering
  \input{../../tables/top_ten.tex}
  \caption{\textit{Top ten career-average adjusted QBR predictions for quarterback prospects in the 2025 draft class. This includes draft-eligible players who played at an NCAA Division I FBS school during the 2024 season.}}
  \label{tab:top-ten}
\end{table}

@@

These four quarterback prospects illustrate the model's interpretability and its alignment with real-world outcomes. Sanders and Ward were both initially expected to go in the first round of the 2025 draft. Sanders' draft position unexpectedly slid, and he ended up being taken in the fifth round in the draft. However, this is likely a result of his refusal to participate in the NFL Scouting Combine, coupled with a reported bad attitude in his interviews, rather than his college performance (\cite{mckenna_what_2025}). In the end, Dart was the other quarterback taken in the first round of the draft.

\begin{table}[H]
  \resizebox{\textwidth}{!}{
    \input{../../tables/side_by_side_similarity.tex}
  }
  \caption{\textit{Top ten similarity scores for each of the top four quarterback prospects in the 2025 NFL draft. The similarity score is the contribution made by each historical prospect to the weighted average which constitutes the reference prospect's prediction in the random forest model.}}
  \label{tab:side-by-side-similarity}
\end{table}

>>>>

\documentclass[Review, sageh, times]{sagej}

@@

All data used for this study are publicly available from Sports Reference (\cite{sports_reference_sports_2025}). Our dataset includes 2,114 quarterbacks who played their final season at the NCAA Division I Football Bowl Subdivision (FBS) between 2001 and 2023 (inclusive) who played more than six total games in their careers and attempted more than five passes per season. The data for each player-season include team, conference, strength of schedule, games played, passing statistics (attempts, completions, touchdowns, interceptions), rushing statistics (attempts, yards, touchdowns), and awards (All-America designation, Heisman voting).

@@

@@

To train the random forest model, we first implemented a random 70\%/30\% train/test split. We chose $B$ = 500 decision trees, each trained on a bootstramp sample of the data, with replacement. To determine the number of variables to consider at each node and the maximum tree size, we used the R package tuneRanger (\cite{probst_tuneranger_2018}), which uses model-based optimization of out-of-bag prediction error to tune the hyperparameters. After 100 iterations of the tuning algorithm, we estimated the optimal number of variables to consider at each node to be 3 and the optimal minimum node size to be 21.

@@

Figure \ref{fig:importance} shows the importance of each feature learned by the random forest model. The model's test RMSE was 8.61 units of QBR, corresponding to explaining 33.3\% of the variance of QBR in the test set. Average yards per season, completions per season and touchdowns per season in college stand out as the most important predictors of NFL success. Note that these three variables are highly correlated with each other---each pairwise correlation is at least 0.94. The next most important variables are final-season Heisman voting, average passing attempts per season and final-season passer rating.

@@

\begin{figure}[H]
    \centering
    \includegraphics[width=0.75\linewidth]{../../figures/3d_plot.pdf}
    \caption{\textit{Predicted adjusted QBR plotted against average passing yards per season and final-season passer rating.}}
    \label{fig:3d-plot}
\end{figure}

@@

\begin{table}[H]
  \centering
  \begin{tabular}{l|r}
    Player & Predicted QBR\\
    \hline
    \input{../../tables/top_ten.tex}
  \end{tabular}
  \caption{\textit{Top ten career-average adjusted QBR predictions for quarterback prospects in the 2025 draft class. This includes only draft-eligible players who played at an NCAA Division I FBS school during the 2024 season.}}
  \label{tab:top-ten}
\end{table}

@@

These four quarterback prospects illustrate the model's interpretability and its alignment with real-world outcomes. Sanders and Ward were both initially expected to go in the first round of the 2025 draft. Sanders' draft position unexpectedly slid, and he ended up being taken in the fifth round in the draft. However, this is likely a result of his refusal to participate in the NFL Scouting Combine, coupled with a reported bad attitude in his interviews, rather than his college performance (\cite{mckenna_what_nodate}). In the end, Dart was the other quarterback taken in the first round of the draft.

\begin{table}[H]
  \resizebox{\textwidth}{!}{
    \begin{tabular}{lr|lr|lr|lr}
      \multicolumn{2}{c|}{\bf Cam Ward} &
        \multicolumn{2}{c|}{\bf Shedeur Sanders} &
        \multicolumn{2}{c|}{\bf Jaxson Dart} &
        \multicolumn{2}{c}{\bf Dillon Gabriel}\\
      Comp & Score & Comp & Score & Comp & Score & Comp & Score\\
      \hline
      \input{../../tables/side_by_side_similarity.tex}
    \end{tabular}
  }
  \caption{\textit{Top ten similarity scores for each of the top four quarterback prospects in the 2025 NFL draft. The similarity score is the contribution made by each historical prospect to the weighted average which constitutes the reference prospect's prediction in the random forest model.}}
  \label{tab:side-by-side-similarity}
\end{table}
